---
title: Simulation
author: Charlotte Wickham
date: Sep 27th 2018
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
```

# Good code

## Pick up from Thursday

---

## Good code

In rough order of importance (for us as statisticians/data scientists):

1. Correct
2. **Obviously** correct: the code is easy to understand
3. Expressive: the code emphasizes the actions not the details
4. Fast

---

## Use existing functions when you can

E.g. `rbinom()` rather than implementing the inverse transform method

* See:
    ```{r, eval = TRUE}
    ?Distributions
    ```
    for those available in the `stats` package.  

* See https://cran.r-project.org/web/views/Distributions.html for packages that implement other distributions.

Inverse transform (Tue) and Rejection method (Lab) are options when nothing exists for the distribution you need.


# Monte Carlo simulations

## Missisippi

Consider the 11 letters in the word *Mississippi*.

```{r}
library(tidyverse)
(mi_letters <- str_split("Mississippi", "")[[1]])
```

What is the probability that no adjacent letters are the same, in a random reordering of the letters?

*Taken from Tijms, Henk. Probability: A Lively Introduction. Cambridge University Press, 2017.*

**Your Turn**: 

* How would you do a simulation to estimate this probability?
* How is this simulation similar to the Russian election example?

---

## General structure of Monte Carlo Simulations

1. Simulate many examples
2. Calculate something on each example
3. Explore the many calculated things

---

## Mississippi

1. Generate random reorderings of the letters in *Mississippi*
2. For each reordering, ask: are all adjacent letters different?
3. Find the proportion of "Yes"s

---

## Russian Election Fraud

1. Generate turnout vectors based on no fraud (Binomial counts based on number on voter list and observed turnout, transformed to turnout percentages)
2. For each turnout vector, count the number of integers.
3. Find the mean and quantiles of the counts

---

## One pretty flexible approach 

First, figure out 1. and 2. on a single example.

Then scale up:

1. Simulate many examples `rerun()`
2. Calculate something on each example `map()`, `map_dbl()`
3. Explore the many calculated things `# depends on goal`

`rerun()` and `map()` are in the purrr package.

---

## One Mississippi

Step 1: How do we get **a** random reordering of the letters in Mississippi?


```{r}
mi_letters
```

---

## `sample(x, size, replace = FALSE)`

> sample takes a sample of the specified size from the elements of x using either with or without replacement.

-- `?sample`

```{r}
(one_reordering <- sample(mi_letters))
```

---

## Two Mississippi

Step 2: Given a reordering, does it have letters next to each other that are the same?

The `rle()` function will be very useful.

**What does it do?**
```{r}
rle(one_reordering)
```
**How can we use it?**

---

## So far

```{r}
one_reordering <- sample(mi_letters) # One example

# Fill in this bit in class

```

---

## Scaling up - many examples with `rerun()`

The first argument is the number of times you'd like to repeat the evaluation of the second argument.

```{r}
many_reorderings <- rerun(10, sample(mi_letters))
```

**Your turn** Take a look at the object `many_reorderings`.  What kind of object is it?  Generate 1000 reorderings instead.

---

## Scaling up - for each example, do something with `map()`

`map()` solves iteration problems, like: for each ___ do ___.

1. First argument is the object you want to iterate over, `many_reorderings`

2. Second argument describes what you want to do. One way, specify a formula (starts with `~`) using `.` as a placeholder for a single example:  `~ any(rle(.)$lengths > 1`

```{r}
map(many_reorderings, ~ all(rle(.x)$lengths == 1))
```

---

## map() returns a list

Use one of its friends instead: `map_dbl()`, `map_lgl()`, `map_int()`, `map_chr()` to get an atomic vector.

**Your turn**: Swap out `map()` for the appropriate function

```{r}
map_lgl(many_reorderings, ~ all(rle(.x)$lengths == 1))
```

---

## All together

```{r}
num_sims <- 1000
many_reorderings <- rerun(num_sims, sample(mi_letters))
adj_letters_same <- map_lgl(many_reorderings, 
  ~ all(rle(.x)$lengths == 1))

# Explore 
adj_letters_same %>% table()
adj_letters_same %>% mean()
```

---

## Your turn

A random sequence of H's and T's is generated by tossing a fair coin $n = 20$ times.  What's the expected length of the longest run of consecutive heads or tails?

*Taken from Tijms, Henk. Probability: A Lively Introduction. Cambridge University Press, 2017.*

---

## Iteration

Common patterns:

1. Do this thing `m` times, `rerun()`
2. Do this thing to each element of `x`, `map()`
3. Do this thing until some condition is satisfied `while`

You can do 1. and 2. with `for` loops, but the purrr functions abstract away the details and let you focus on *"this thing"*.

*(You also don't run the risk of writing an inefficient `for` loop)*

**But remember** R loves working with vectors. Don't iterate over the elements of a vector, when a function exists to handle the whole vector.

--- 

## Questions

* How many simulations do we need? I.e. what should `num_sims` be?
* What do we do when the simulation part is complicated?  What do we do when the calculation part is complicated?


