---
title: C++ or Parallelization
author: Charlotte Wickham
date: Nov 6th 2018
---

```{r setup, message = FALSE}
library(tidyverse)
library(here)
set.seed(18973)
```

# C++

A lower level compiled language which means a much lower overhead on `for` loops.

Writing a function in C++ may bring big speed improvements for things like:

* Loops that aren't easily vectorised
* Problems that involve calling functions millions of times
* Advanced data structures or algorithms R doesn't provide

We don't have time to learn whole new language, but I want to expose you to a little C++, so you know how to get started and can see the benefits.

Mostly you would use C++ to write small self contained functions, that are called from your R scripts.  I.e. rewrite just a few functions that are called often (or are inefficient) not the whole simulation/analysis.

Resource: https://adv-r.hadley.nz/rcpp.html

## Rcpp

An R package that makes it easy to connect a C++ definition of a function to R:
```{r}
library(Rcpp)
```

**On your own machine**: You'll need a working C++ compiler, see 
https://adv-r.hadley.nz/rcpp.html#prerequistes

## Spot the difference

Here are two functions that do the same thing, one implemented in R and one implemented in C++.

In R:
```r
add <- function(x, y, z){
  x + y + z
}
```

In C++:
```cpp
int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}
```

**Your Turn** Brainstorm the ways the two function definitions are different.

## R vs. C++

Differences:

* Defining a function in C++: define it like you call it 
    ```
    function_name(arg1_name, arg2_name){
      /* Body of function */
    }
    ```
* Explicit types for both function output, arguments and any intermediate objects

* Finish lines with `;`

* Explicit `return()` statement

* Use `=` not `<-`


## Using a C++ function in R

Three possibilities:

1. Pass the definition inside `'` to `cppFunction()`:
    ```{r}
    cppFunction('int add(int x, int y, int z) {
      int sum = x + y + z;
      return sum;
    }')
    ```

2. Put the definition (plus some additional lines) into a special `Rcpp` code chunk in your RMarkdown (look at the RMD version of these notes to see this in action):
    ```{Rcpp}
    #include <Rcpp.h>
    using namespace Rcpp;
    
    // [[Rcpp::export]]
    int add(int x, int y, int z) {
      int sum = x + y + z;
      return sum;
    }
    ```
          

3. Put the definition (and the same additional lines) into their own file, e.g. `file.cpp` and use `sourceCpp()`:
    ```{r, eval = FALSE}
    library(Rcpp)
    sourceCpp("functions.cpp")
    ```

Either way, then the function, `add()` in this case, can be used just like any R function:
```{r}
add
```

```{r}
add(1, 2, 3)
```

C+ is strict on inputs, although coercion will happen automatically
```{r}
add(1.3, 2, 3) # 1.2 gets coerced to integer (rounded down)
```
```{r}
add(c(1, 2), 1, 2) # vector input raised an error
```


## Types in C++

In C++ scalars (single numbers) are a different type to vectors.

C++ types:

|        | integer         |  double          | character          | logical         |
|--------|-----------------|------------------|--------------------|-----------------|
| Scalar | `int`           | `double`         | `String`           | `bool`          |
| Vector | `IntegerVector` | `NumericVector`  | `CharacterVector`  | `LogicalVector` |
| Matrix | `IntegerMatrix` | `NumericMatrix`  | `CharacterMatrix`  | `LogicalMatrix` |

## Another spot the difference

`signR()` defined in R:
```{r}
signR <- function(x) {
  if (x > 0) {
    1
  } else if (x == 0) {
    0
  } else {
    -1
  }
}
```

`signC()` defined in C++:
```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
int signC(int x) {
  if (x > 0) {
    return 1;
  } else if (x == 0) {
    return 0;
  } else {
    return -1;
  }
}
```

**Your Turn** How are `if` statements different in C++?

There isn't any!

## A function with a `for` loop

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double sumC(NumericVector x) {
  int n = x.size(); 
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}
```

In C++:

* you call a method with `object_name.method_name()`, e.g. `x.size()` calls the `size()` method associated with the object `x` - in this case the length of the vector.

* `for` loops look like `for(init; check; increment)`:

    * `init` the initial value(s) `int i = 1`, remembering to define its type.
    * `check` the value against a stopping rule `i < n`
    * `increment` the value for the next iteration `++i` means `i = i + 1`

* **Vector indicies start at 0**. **VECTOR INDICIES START AT 0!**

* There are modify in place operations `total += x[i]` is equivalent to `total = total + x[i]`

## Your Turn

Can you figure out what this function does?  What is the equivalent in base R?

```{Rcpp, eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector f2(NumericVector x) {
  int n = x.size();
  NumericVector out(n);

  out[0] = x[0];
  for(int i = 1; i < n; ++i) {
    out[i] = out[i - 1] + x[i];
  }
  return out;
}
```


```{r}
f2(c(1, 3, 4, 2, 1))
```

## Speed 

Remember the `abs()` example from last week:

```{r}
abs_loop <- function(vec){
  for (i in 1:length(vec)) {
    if (vec[i] < 0) {
      vec[i] <- -vec[i]
    }
  }
  vec
}

abs_sets <- function(vec){
  negs <- vec < 0
  vec[negs] <- -vec[negs] 
  vec
}
```

Here's a C++ version:
```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector absC(NumericVector x){
  int n = x.size();
  for(int i = 0; i < n; i++){
    if(x[i] < 0){
      x[i] *= -1;
    } 
  }
  return x;
}
```

```{r}
x <- -50:50
bench::mark(
  abs_loop = abs_loop(x),
  abs_sets = abs_sets(x),
  absC = absC(x),
  abs_base = abs(x))
```

Despite operating one element at a time, the C++ version is about as fast as the set version.  Still not as fast as the base version.

## Rcpp sugar

A part of the Rcpp package to make translating code easier. 

*Rcpp sugar* "bring[s] a subset of the high-level R syntax in C++."

Arithmetic and logical operators are vectorized

Access to common vectorized functions in R like: `any()`, `ifelse()`, `is.na()`

Access to common distribution functions

## C++ STL

A software library for the C++ with common algorithms and data structures.

E.g. `std::max()`,  `std::random_shuffle()`, `std::partial_sort()`

Back at 9:00

# Parallelization

**Embarrasingly** parallel problems

> is one where little or no effort is needed to separate the problem into a number of parallel tasks

-- https://en.wikipedia.org/wiki/Embarrassingly_parallel

Simulation studies often fall into this category: 

* simulating 100 processes, in 10 parallel tasks is the same as simulating 1000 processes in 1 task.
* or applying a method to 1000 simulated data sets, could be done in 10 parallel tasks each applying the method to a 100 simulated data sets.

## Quick talk about multiple cores, multiple processors and clusters

What do they share?

On my machine
```{r}
parallel::detectCores()
```

## Two parts to running in parallel

*(Charlotte's rough conceptual breakdown)*

**Setup** specify what parallel means on your set up, i.e. use multiple cores, background sessions, remote computers, a cluster etc.

**Execute** changes to your code that make use of the parallel set up. (Technically, explicit parallelism as opposed to implicit parallelism).

Lot's of R packages available: https://cran.r-project.org/web/views/HighPerformanceComputing.html

## An example package: furrr

**Your turn**: Start reading the readme at: https://github.com/DavisVaughan/furrr

What is the goal of the package? What packages does it build on?

## Execution

**Your turn**: Scroll down to the first example.  What does it illustrate?

## Setup

**Your turn**: Keep reading the examples, what changes to get the map operation to run in parallel? How does the author provide proof that it is running in parallel?

## On your machine?

**Your turn**: Try running the example on your own, can you replicate the timings?













 