---
title: S3 Object Oriented Programming in R
author: Charlotte Wickham
date: Nov 8th 2018
---

```{r setup, message = FALSE}
library(tidyverse)
library(here)
library(sloop)
set.seed(36528)
```

# Motivation

Some code from last week:
```{r}
# some fake data
true_pi <- 0.8
true_mu <- c(0,   2)
true_sigma <- c(1,   0.5)
n <- 250
pop <- rbinom(n, size = 1, prob = 1 - true_pi) + 1
x <- rnorm(n, mean = true_mu[pop], sd = true_sigma[pop])

dmix <- function(x, theta){
  pi <- theta[1]
  mu_1 <- theta[2]
  mu_2 <- theta[3]
  sigma_1 <- theta[4]
  sigma_2 <- theta[5]
  pi * dnorm(x, mean = mu_1, sd = sigma_1) +
   (1 - pi) * dnorm(x, mean = mu_2, sd = sigma_2) 
}
nllhood_mix <- function(theta, x){
  d <- dmix(x, theta)
  -1 * sum(log(d))
}
```

Recall:

```{r, warning = FALSE}
mle1 <- optim(par = c(1, 0, 1.8, 1.2, 0.7), fn = nllhood_mix, x = x, method = "BFGS")
mle1
```

```{r, warning = FALSE}
mle2 <- optim(par = c(0.5, 0, 1, 2, 1), fn = nllhood_mix, x = x, method = "BFGS")
mle2
```

**Your Turn** What do you do with this output?  How might you improve what the output looks like to help you as a user?

**Our goal**: To get a more user friendly output from `optim()`


# Object Oriented Programming (OOP)

A programming paradigm where objects are central.  Two important terms:

* **Class** defines what an object is

* **Methods** describe what an object can do

Gives us a way to have the same function operate differently on different classes of object, e.g. `summary()`

```{r}
z <- factor(sample(letters[1:5], size = 10, replace = TRUE))
y <- rnorm(n = 10)
summary(z)
summary(y)
```

There is more than one OOP system in R, we'll focus on S3.

## S3

The most common OO system R. Also the simplest.

> S3 allows your functions to return rich results with user-friendly display and programmer-friendly internals. S3 is used throughout base R, so it's important to master if you want to extend base R functions to work with new types of input.

-- [Advanced R](https://adv-r.hadley.nz/oo.html)

## Why bother?

**Primarily** to make your functions easier to use (for you or others), by creating a new class and defining some methods:

* Have your functions print nice output
* Implement methods for familiar functions, e.g. `plot()`, `summary()`

**More advanced** make it easy for others to extend your work for other data types, by supplying generic functions.

## Orientation

As our functions get more complicated they need to return complicated things.

E.g. Linear models

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
str(mod)
```

Complicated objects are built on top of simple objects (vectors, matrices, lists etc.) with the additional of attributes.

E.g. `lm()` returns a list along with the a `class` attribute.

The `class` attribute is special, it's how R knows what kind of object something is.

Use `class()` to see the class, 
```{r}
class(mod) # the class of this object
```
`typeof()` to see the base type the object is built on, 
```{r}
typeof(mod) # the base type this object is built on
```
and `attributes()` to see all attributes:
```{r}
attributes(mod)
```

**Your turn** What class is `z`? What base object is it built on? What other attributes does this object have?

```{r}
z <- factor(sample(letters[1:5], size = 10, replace = TRUE))

class(z)
typeof(z)
attributes(z)
```

Does everything have a class?  `class()` will return something, even if the object doesn't have a `class` attribute:
```{r}
class(y)
attributes(y)
```

## Methods

In S3 methods are associated with generic functions (generics for short).

This is different to most other OO languages out there, which use encapsulated OO - the methods are associated with the object.

When an S3 generic function is called, it looks at the class of the object passed to it, and searches for the appropriate method, of the form `generic.class`, e.g. 
```{r}
summary(mod) # calls summary.lm
class(mod)
```
looks for finds and calls `summary.lm()`

This process is known as method dispatch.  You can see the process with `s3_dispatch()` in the `sloop` package:
```{r}
s3_dispatch(summary(mod))
```

`=>` the method that is called, `*` a method that is defined but not called.
  
If you want to know if a function is a generic you can use `ftype()` in sloop
```{r}
ftype(summary)
ftype(summary.lm)
```
  
**Your turn**  You can use `.` in a name in R without it being a method.  Can you figure out if `t.test()` is the `t` method for objects of class `test`?  What about `t.data.frame()`?

```{r}
# install.packages("sloop")
library(sloop)

ftype(t.test) # S3 generic, not a method for `t()` for class "test"
ftype(t.data.frame) # S3 method for `t()` for class "data.frame"
```

```{r}
fake <- data.frame(x = 1:10)
class(fake)
s3_dispatch(t(fake))  # let's you see how R found the method
```

Unanswered question: What if we define a `frame` class and a `t.data()` generic, then and a `frame` method for `t.data()`, do things break?

## `print()` is a generic

```{r}
ftype(print)
```

It gets called when you ask for an object
```{r}
mod # is the same as
print(mod)
```

Behind the scenes it finds the `print.lm()` function
```{r}
s3_dispatch(print(mod))
```

## `class` doesn't get any special treatment

You can set the class of an object just like any usual assignment:
```{r}
class(mod) <- "table"
mod
```

But you probably wouldn't want to...

If we remove the class of `mod`
```{r}
class(mod) <- NULL
mod 
s3_dispatch(print(mod))
```

It just prints like any list.  This isn't very useful output...why?

## To build a new object

You can either make the object from it's base type and add a class (or any other attributes):
```{r}
y <- 1:6
class(y) <- "my_special_kind_of_vector"
attr(y, "max") <- 6
y
```

Or use `structure()` to do it in one go
```{r}
y <- structure(1:6, max = 6, class = "my_special_kind_of_vector")
y
```

# Back to our goal

Let's build a new class to hold `optim()` output, then write a `print()` method that prints output in more useful fashion.

General procedure:

1. Decide on a name for the class
2. Decide what structure/properties the class should have
3. Write a low level *constructor* function creates this kind of object, and a user friendly *helper* function to easily get an example.
4. Write methods

**Your turn** What shall we call the class to hold an `optim()` result?

1.  `my_new_class`, `optim`, `my_optim`, `cl_optim`, `optim_class`, `optim_out`, **`optimum`**

2. Takes the same structure as existing `optim()` output - list with certain elements.  We aren't going to check for these elements in our constructor.  

## A constructor

https://adv-r.hadley.nz/s3.html#s3-constrcutor

First we'll write a function that creates an object of class `optimum`.  

By convention, call it `new_classname()`:
```{r}
new_optimum <- function(l){
  # This is where you need to enforce any desired structure
  stopifnot(is.list(l))
  
  structure(
    l,
    class = "optimum"
  )
}

new_optimum(list(par = c(1, 2, 3), y = "weird"))
```

I'm being a little lazy checking the input, but I'll assume the incoming list has all the components of an `optim()` output.

```{r}
new_optimum(mle1)
s3_dispatch(print(new_optimum(mle1)))
```

## A helper

https://adv-r.hadley.nz/s3.html#helpers

It's useful to have a function `class_name()` that conveniently creates object of this class, which we'll assume just passes it's arguments onto `optim()`:
```{r}
optimum <- function(..., quiet = TRUE){
  if(quiet){
    suppressWarnings({
      opt <- optim(...)
      })
  } else {
    opt <- optim(...)
  }
  new_optimum(opt)
}
```

```{r}
optimum(par = c(1, 0, 1.8, 1.2, 0.7), 
  fn = nllhood_mix, x = x, method = "BFGS")  
```

## Our first method `print.optimum()`

To define a new method, you just write a regular function with a special name `generic.class()`:
```{r}
print.optimum <- function(x, ...){
  cat("Final parameter values: \n")
  cat(x$par)
  invisible(x)
}

optimum(par = c(1, 0, 1.8, 1.2, 0.7), 
  fn = nllhood_mix, x = x, method = "BFGS")  
```

**Your turn** Edit the method to just print the final parameter values.

## Better output

We should check that value of `convergence` and print a message if `optim()` didn't converge: 
```{r}
print.optimum <- function(x, ...){
  if(x$convergence != 0){
    cat("Failure to converge \n")
  } else {
    cat("Coverged! \n")
  }
  cat("Final parameter values \n")
  print(x$par)
  invisible(x)
}
optimum(par = c(1, 0, 1.8, 1.2, 0.7), 
  fn = nllhood_mix, x = x, method = "BFGS")  
optimum(par = c(0.5, 0, 1, 2, 1), fn = nllhood_mix, x = x, method = "BFGS")
```


Even better we might check some more convergence values and add some color! (might need `install.packages("crayon")`):
```{r}
print.optimum <- function(x, ...){
  if(x$convergence != 0){
    cat(crayon::red("Failure to converge \n"))
    if (x$convergence == 1){
      cat("Iteration limit `maxit` reached. \n")
    } else {
      cat(x$message)
    }
  } else {
    cat(crayon::green("Coverged! \n"))
  }
  cat("Final parameter values \n")
  print(x$par)
  invisible(x)
}
optimum(par = c(1, 0, 1.8, 1.2, 0.7), 
  fn = nllhood_mix, x = x, method = "BFGS")  
optimum(par = c(0.5, 0, 1, 2, 1), fn = nllhood_mix, x = x, method = "BFGS")
```

## Another method

**Your turn** Is `coef()` a generic?  Write a `coef()` method for `optimum` that returns the parameter estimates.

```{r}
ftype(coef) # Yes, it's a generic
coef # Our method needs to accept the same arguments as the generic

coef.optimum <- function(object, ...) {
  object$par
}

mle_1a <- optimum(par = c(1, 0, 1.8, 1.2, 0.7), 
  fn = nllhood_mix, x = x, method = "BFGS")  

coef(mle_1a)
```


## FYI stats4::mle()

Does a similar thing to what we have implemented but it uses the S4 OO system.
