---
title: Documentation and Testing 
author: Charlotte Wickham
date: Nov 15th 2018
---

```{r setup, message = FALSE}
library(tidyverse)
set.seed(197387)
knitr::opts_chunk$set(eval = FALSE)
```

[Presentation Schedule](https://github.com/ST541-Fall2018/presentations)

# Documentation and Testing

Useful techniques that come from R package development.  

* **Testing** - formal tests of functions to ensure correctness as changes are made

* **Documentation** - formal R documentation so others (and future you) know how to use functions

Adpated from *Tidy Tools* by Hadley Wickham CC BY NC 4.0.

# Testing

**Resource**:  http://r-pkgs.had.co.nz/tests.html

You will need this [RStudio Cloud Project](https://rstudio.cloud/spaces/4116/project/136930)

## Motivation

We want a function that adds columns from one dataframe to another at any position.

*E.g Give example on board*

In `R/add_cols.R` a first attempt `add_cols()`, ignore comments for now.

To load this function:
```{r}
devtools::load_all()
```

Or Crtl/Cmd + Shift + L

**Your turn** Come up with a simple example that we can test the `add_cols()` with.  

What does the input to `add_cols()` look like?  What should the output look like?  

Try it out in a new R script or Rmarkdown.

**What's wrong with the current version of `add_cols()`?**  Don't fix it yet!

Testing is a way to formalise this process.

## Set up testing infrastructure

With `add_cols.R` open:
```{r}
usethis::use_test()
```

Replace this code:
```{r}
test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})
```

With this code:
```{r}
test_that("can add column at any position", {
  df1 <- data.frame(a = 3, b = 4, c = 5)
  df2 <- data.frame(X = 1, Y = 2)
  at_pos <- function(i) {
    add_cols(df1, df2, where = i)
  }
  expect_named(at_pos(1), c("X", "Y", "a", "b", "c"))
  expect_named(at_pos(2), c("a", "X", "Y", "b", "c"))
  expect_named(at_pos(3), c("a", "b", "X", "Y", "c"))
  expect_named(at_pos(4), c("a", "b", "c", "X", "Y"))
})
```

## Test 

**Your turn** 

Run
```{r}
devtools::test()
```
Or Crtl/Cmd + Shift + T to run the tests.

Verify that they fail, then try to fix `add_cols()` to fix it.  

## Testing workflow

1. Write code
2. Test code: `devtools::test()` or Crtl/Cmd + Shift +
3. Repeat

Some other uses:

* When you stop work leave a test failing, that way you know where to start when you come back.
* Test driven development: start by writing the tests before the function.

## Writing tests

Now we'll add another function `add_col()` in it's own R file.

`add_col(x, name, value, where)` adds a single new column - **draw picture on board**.

Create the file in the right place:
```{r}
usethis::use_r("add-col")
```

Copy in this code:
```{r}
add_col <- function(x, name, value, where) {
  df <- setNames(data.frame(value), name)
  add_cols(x, df, where = where)
}
```

```{r}
devtools::load_all()
```

Then run:
```{r}
usethis::use_test()
```
to open up the appropriate file for the tests.

## Some useful testing functions

`expect_equal(a, b)` `TRUE` if `a` and `b` are that same based on `all.equal()`. 

**Your turn**

Add the following test, filling in the blank (`___`) to test the output.

```{r}
test_that("where controls position", {
  df <- data.frame(x = 1)
  expect_equal(add_col(df, "y", 2, where = 1), ___)
})
```

Add another test for `add_col(df, "y", 2, where = 2)`

`expect_error(a, "b")` checks `a` results in an error message with a message that contain `"b"`.

**Your turn** What should happen in this case? 

```{r}
add_col(df, "y", 2, where = "a")
```

# Documentation

**Resource**:  http://r-pkgs.had.co.nz/man.html

Special comments where the function is defined get turned into documentation, e.g. in `add_cols.R`

```
#' Add a column to a data frame
#'
#' Allows you to specify the position. Will replace existing variable
#' with the same name if present.
#'
#' @param x A data frame
#' @param name Name of variable to create. If a variable of that name
#'   already exists it will be replaced
#' @param value Values to insert.
#' @param where Position to insert. Use 1 to insert on LHS, or -1 to insert on
#'   RHS.
#' @examples
#' df <- data.frame(x = 1:5)
#' add_col(df, "y", runif(5))
#' add_col(df, "y", runif(5), where = 1)
#'
#' add_col(df, "x", 5:1)
```

## Documentation using roxygen2

Workflow:

1. You write specially formatted comment above function definitions
2. `devtools::document()` generates `.Rd` files
3. Preview, `?topicname` R converts `.Rd` to HTML for viewing.

**Your turn** 

```{r}
devtools::document()
?add_cols
```

Fix the typos in `add_cols()` documentation, re-document to check your work.

Typos:

* "Another" in the title is spelled wrong.
* The "x" in the description of the `where` argument should look like code.
* In Details, "right" should be "left"

## Starting from scratch

For a function with no documentation.  Put your cursor in the body of the function: Code -> Insert Roxygen Skeleton

Try it with `add_col()`.

## Next week

* Presentations
* Package/Project level documentation
