---
title: Lab 02
author: Charlotte Wickham
date: Sep 27th 2018
---
  
## Learning Objectives

In class we saw one method for turning a sample from a Uniform(0, 1) random varible into a sample from some other distibrution: the inverse transform method.

In this lab you'll see another technique for turning a sample from one distribution into a sample from another: the rejection method.

By the end of this lab you should be able to:

* Describe the steps in Rejection sampling
* Given a proposal and target density choose a value for $c$
* Describe how the choice of $c$ affects the efficiency of the sampling
* Identify the pattern of iteration suitbale for a `while` loop
  
If you want more details see Section 5.2 in Ross in the readings for the week.
  
```{r, message = FALSE}
library(tidyverse)
```

## Acceptance Rejection Algorithm

**General Idea** 

If you can simulate from distribution with probability density function (pdf) $g(x)$, then you can sample from a distribution with pdf $f(x)$ by sampling $Y$ from $g(x)$ and accepting the sample with probability proportional to $\frac{f(Y)}{g(Y)}$. 

**The Intuition**

When we sample from $g(x)$, also known as a proposal distribution, in some regions we'll get more observations that we should expect from $f(x)$, our target distribution, so to decrease their prevalance we accept fewer of them, i.e. we accept them with some value less than 100%.  

For example, if we draw $Y$ from $g(Y)$, and the ratio $\tfrac{f(Y)}{g(Y)} = \tfrac{1}{3}$, then it's 3 times less likely to see this value from our desired distribution than our target distibrution, and we should only accept it with probability 1/3 to account for that.  

Of course there will be other draws of $Y$, where $\tfrac{f(Y)}{g(Y)} > 1$, but we can't accept something with a probability above 1!  So, in practice we find a value $c$ such that $\tfrac{f(Y)}{g(Y)} \le c$ for all possible $Y$, and instead accept draws with probability $\tfrac{f(Y)}{cg(Y)}$.

*Sam will talk through this*

**The Algorithm**

If you want to sample $X$ from $f(x)$, based on draws from $g(x)$, first find a constant $c$ such that
$$
\frac{f(x)}{g(x)} \le c \quad \text{for all } x
$$
Then:

1. Draw a sample $Y$ from $f(x)$
2. Accept $Y$ with probability $\tfrac{f(Y)}{cg(Y)}$.  That is, draw U from Unif(0, 1) and if $U < \tfrac{f(Y)}{cg(Y)}$ set $X = Y$, otherwise return to step 1.


## An example

To see this in action you'll work through example 5d in Ross. 

In this example, we want a sample from a distribution with p.d.f.
$$
f(x) = 20 x (1 - x)^3, \quad 0 < x < 1
$$

But what we have at hand, is a way to sample from the Uniform(0, 1), 
$$
g(x) = 1, \quad 0 < x < 1
$$

## Find a constant

Let's take a look at the target density.  To do so, we'll create a function for $f(x)$:
```{r}
f_x <- function(x){
  20 * x * (1 - x)^3
}
```

And then plot it:
```{r}
ggplot(data.frame(x = c(0, 1)), aes(x)) +
  stat_function(fun = f_x, geom = "line") +
  labs(y = "f(x)", x = "x")
```

**What would be a good value of $c$?** I.e. what value for $c$ will ensure our target density is always below $cg(x)$?

Use the figure to pick $c$ and set it below:
```{r}
constant <- 2.5  # my choice, yours might be different
```

## Generating a sample

We first sample $Y$ from our proposal density, Uniform(0,1)
```{r}
set.seed(7416) # So your results will match those online
Y <- runif(1)
```

Then randomly accept this value with probability $\tfrac{f(Y)}{cg(Y)}$
```{r}
U <- runif(1)
(accept <- U < (f_x(Y)/(constant * dunif(Y))))
```
I got `FALSE`, i.e. I shouldn't accept this number (if your `constant` is very different from mine, you might get a different result).

In which case I should try again:
```{r}
Y <- runif(1)
U <- runif(1)
(accept <- U < (f_x(Y)/(constant * 1)))
```

Doh, `FALSE` again. 

```{r}
Y <- runif(1)
U <- runif(1)
(accept <- U < (f_x(Y)/(constant * 1)))
```
Finally, so this $Y$ is my draw
```{r}
Y
```
You can show the expected number of draws before accepting a value is $c$, so it isn't surprising it took me three times here.

**Your Turn**

* Ross finds $c =  135/64$.  What's the advantage of using calculus to find the smallest $c$ that satifies the requirement? 

* The target distribution and the proposal distribution in this example have the same support?  What would happen if for some $x$, $g(x) > 0$ and $f(x) = 0$?  Or vice versa, for some $x$, $g(x) = 0$ and $f(x) > 0$

## Automating

If we want to automate drawing a single sample (no matter how many rejections it takes), we would need to use a `while()` loop.  The body of a `while()` loop will continue to be evaluated until the condition (inside the `()`) evaluates to `TRUE`.

As a simple example, we could keep adding 1 to `i` until we hit 5:
```{r}
i <- 0
while(i < 5){
  cat(i, "\n")
  i <- i + 1
}
```
Or in words: set `i` to be zero, then while `i` is less than five, tell me what `i` is and add one.

We might implement the rejection algorithm for our example like this:
```{r}
accept <- FALSE
while(!accept){
  Y <- runif(1)
  U <- runif(1)
  accept <- U < f_x(Y)/constant
}
Y
```

Of course to get another number we need to rerun the entire block of code.  A better approach might be to wrap this chunk into a function:
```{r}
draw_one <- function(quiet = FALSE){
  accept <- FALSE
  while(!accept){
    Y <- runif(1)
    U <- runif(1)
    accept <- U < f_x(Y)/constant
  }
  Y
}
```

Now we can easily run it many times:
```{r}
many_draws <- rerun(1000, draw_one()) %>% 
  flatten_dbl() # to get a vector rather than a list
```

And check our draws look about right:
```{r}
ggplot(mapping = aes(x = many_draws)) + 
  geom_histogram(aes(y = stat(density))) +
  stat_function(fun = f_x, geom = "line") +
  labs(y = "f(x)", x = "x")
```

Looking good!

**Your Turn**:

* This function has a big problem in that it relies on objects defined outside of the function.  What are these objects? Which objects should be passed in as arguments?
* What needs to change in this function if we use a different proposal density? 
* **Challenge** A more efficient method for drawing sample of size $n$ might be to draw many $Y$ and $U$ in a chunk, keep all those that meet the acceptance criterion, then generate a new chunk if we don't have $n$ yet.  Try implementing this method.


You might like to compare this version of `draw_one()` to the one above.  What is different? 

```{r}
draw_one <- function(target_density, proposal_density, proposal_sampler, constant){
  accept <- FALSE
  while(!accept){
    Y <- proposal_sampler(1)
    U <- runif(1)
    accept <- U < target_density(Y)/(constant*proposal_density(Y))
  }
  Y
}

# our example
draw_one(target_density = f_x, 
  proposal_density = dunif, 
  proposal_sampler = runif, 
  constant = 2.5)
```
