---
title: Lab 08 - S3 OO programming in R
author: Charlotte Wickham
date: Nov 9th 2018
---

```{r setup, message=FALSE}
library(tidyverse)
set.seed(181763)
```

Get ready to play the slots! In this lab you'll follow along with the chapter on S3 programming in [Hands-On Programming with R](https://rstudio-education.github.io/hopr/s3.html) (HOPR).

To get oriented start by reading the [introduction to the slot machine project](https://rstudio-education.github.io/hopr/project-3-slot-machine.html).

## Introduction

A slot machine randomly draws three symbols, and depending on the combination of symbols that are drawn, a player may receive a payout. In the slot machine implemented in HOPR the symbols that can be drawn and their probabilities of being drawn are captured in the function `get_symbols()`:
```{r}
# from https://github.com/rstudio-education/hopr
get_symbols <- function() {
  wheel <- c("DD", "7", "BBB", "BB", "B", "C", "0")
  sample(wheel, size = 3, replace = TRUE, 
    prob = c(0.03, 0.03, 0.06, 0.1, 0.25, 0.01, 0.52))
}
```
where `DD` is diamonds, `7` sevens, `BBB` triple bars, `BB` double bars, `B` a single bar, `C` cherries and `0` zeros.

You can run `get_symbols()` to see an example of the symbols drawn on a play of the machine:
```{r}
get_symbols()
```

The payout is complicated for the particular type of machine being emulated (you can see the details in Table 9 in [HOPR](https://rstudio-education.github.io/hopr/programs.html)) and is captured in the `score()` function:
```{r}
score <- function (symbols) {
  # identify case
  same <- symbols[1] == symbols[2] && symbols[2] == symbols[3]
  bars <- symbols %in% c("B", "BB", "BBB")
  
  # get prize
  if (same) {
    payouts <- c("DD" = 100, "7" = 80, "BBB" = 40, "BB" = 25, 
      "B" = 10, "C" = 10, "0" = 0)
    prize <- unname(payouts[symbols[1]])
  } else if (all(bars)) {
    prize <- 5
  } else {
    cherries <- sum(symbols == "C")
    prize <- c(0, 2, 5)[cherries + 1]
  }
  
  # adjust for diamonds
  diamonds <- sum(symbols == "DD")
  prize * 2 ^ diamonds
}
```

This can be used to figure out the payout for a particular symbol set:
```{r}
(one_play <- get_symbols())
score(one_play)
```

Finally, `play()` combines the two steps: drawing and scoring
```{r}
play <- function() {
  symbols <- get_symbols()
  print(symbols)
  score(symbols)
}
```

```{r}
play()
```

## Reading S3 Chapter

OK, now you have the setup you can start with the S3 section: https://rstudio-education.github.io/hopr/s3.html

Here are some questions labelled with corresponding section number that you should answer after reading and working through each section.

### 10 S3

* What's the difference between the current output of `play()` and the desired output of `play()`
* Examine the object `one_play` (created in this section), what is missing from it.

### 10.2 Attributes

This section refers to the variable `DECK` which you can get with
```{r, results = "hide"}
DECK <- read.csv(url("https://gist.githubusercontent.com/garrettgman/9629323/raw/ee5dfc039fd581cb467cc69c226ea2524913c3d8/deck.csv"))
DECK
```


* Do Exercise 10.1, which problem from above does this solve?

* How does `slot_display()` go part-way to solving the other problem?  Why doesn't it completely solve the problem?

### 10.3 Generic Functions

* How do you see all the methods associated with a function? E.g. what methods are available for `summary()`?

### 10.4 Methods

* What happens when a object has more than one class?

* Do exercise 10.2

### 10.5 Classes

* How do you see all the methods associated with a class?

## Going further

By the end of the chapter, `play()` has been altered to always return an object with class `slots`:
```{r}
play <- function() {
  symbols <- get_symbols()
  structure(score(symbols), symbols = symbols, class = "slots")
}
```

And a `print()` method for `slots` is defined relying on a function called  `slot_display()`:
```{r}
slot_display <- function(prize){

  # extract symbols
  symbols <- attr(prize, "symbols")

  # collapse symbols into single string
  symbols <- paste(symbols, collapse = " ")

  # combine symbol with prize as a regular expression
  # \n is regular expression for new line (i.e. return or enter)
  string <- paste(symbols, prize, sep = "\n$")

  # display regular expression in console without quotes
  cat(string)
}

print.slots <- function(x, ...) {
  slot_display(x)
}
```

In class we saw it was best practice to create both a constructor and a helper function for a new class, so let's alter the above to do that.

[Advanced R](https://adv-r.hadley.nz/s3.html#s3-constrcutor) recommends our **constructor** function should:

* Be called `new_myClass()` where we substitute `myClass` for our class name.  So in this case `new_slots()`
* Have one argument for the base object, and one for each attribute. In this case `score` is the base object and `symbols` is an attribute.
* Check the type of the base object and the types of each attribute. `score` should be a length one numeric value, and `symbols` a length 3 character vector.

Constructors will always finish with a call to `structure()` and set the appropriate class. 

**Fill in the following function to complete the constructor**
```{r}
new_slots <- function(){
  
  
}
```

Rewrite `play()` to make use of the new constructor function
```{r}
play <- function() {
  symbols <- get_symbols()
  structure(score(symbols), symbols = symbols, class = "slots")
}
```


It may also be useful to have a **helper** function, that according to Advanced R should:

* Have the same name as the class, e.g. `myclass()`.

* Finish by calling the constructor, and the validator, if it exists.

* Create carefully crafted error messages tailored towards an end-user.

* Have a thoughtfully crafted user interface with carefully chosen default values and useful conversions.

This would allow us to quickly make a `slots` object.  For example we might want to create specific instances of slot machine output for testing, in which case we might expect to be able to do something like:

```r
slots(c("BB", "B", "DD"))
# BB B DD
# $0
```

Notice a useful default might be that if the score isn't supplied `score()` is used to find it, one way to do this is by providing a default value of `NULL` and checking for it in the body of the function.

**Create a helper function for `slots()`**, it should at least:

* check `symbols` is of length 3
* if `score` is `NULL` (use `is.null()` to check), calculate it with the `score` function 
* call the constructor function at the end

```{r}
slots <- function(symbols, score = NULL){
  
}
slots(c("BB", "B", "DD"))
```

Once you have a helper consider the following two possible results:
```{r, results = "hide"}
slots(c("BB", "B", "DD"))
slots(c("BB", "B", "BBB"))
```

It might be nice if we win any money that a congratulations message appears.  

**Edit the print method to make this change**. Bonus points if it prints in color.

## Other possible extensions

* What else should be checked if someone constructs their own `slots` output with `slots()`?  Where is the best place to make the checks?

* What other methods might be useful for the `slots` class?

* Use the `emo` package to add some more exciting symbols, when printing:
    ```{r}
    # devtools::install_github("hadley/emo")
    emo::ji("diamonds")
    emo::ji("cherries")
    emo::ji("seven")
    ```

*  When being using interactively (see `interactive()` to check), put a delay between each symbol being printed using `sys.sleep()`, and for the payout to be displayed.

